---
title: "Analysis of registered covid-19 cases and numbers of death"
author: "Fanni Kiss"
date: '2020 11 29 '
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 

```

## Introduction

Current report aims to analyse the pattern of association between registered COVID-19 cases per capita and registered numbers of death per capita due to COVID-19 on 22 September 2020.
GitHub <https://github.com/fanni-k/DA2>.


```{r include = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
# Calling the cleaned data from my computer
my_path <- "C:/Users/Fandurka/Google Drive/CEU/DA2/assignment_1/data/"
df <- read_csv(paste0(my_path,'clean/covid_pop_09_22_2020_clean_3.csv'))
```

## Exploring the data

The distribution of each variables are all skewed with a long right tale. In the summary table (Appendix, Teble 1), we can observe that there are some extreme values, which are presumably not measurement errors. Thus, extreme values are involved into the further analysis. 
The analysis focuses on two variables: the number of cases per capita (dependent variable) and the number of death per capita (explanatory variable). Both of them skewed with a long right tale and contains only positive numbers.

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', fig.show='hide'}
####
# 
# Checking all variables on histograms and summary statistics
library(magrittr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(knitr)
        
df %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~key, scales = "free") +
  geom_histogram()+
  theme_wsj() + 
  scale_fill_wsj()

kable(summary( df ) , caption = "Descriptive statistics of the variables")

##
# Creating two new variables:
#   1. number of registered death per capita (y variable)
#   2. number of registered cases per capita (x variable)
# Note: the ratios are very low values, so they are multiplied by 10000 for the proper scaling
df <- df %>% mutate( death_per_capita = death/population)
df <- df %>% mutate( case_per_capita = confirmed/population)


# Checking the distribution of x variable
ggplot(data = df , aes(x = death_per_capita)) +
  geom_histogram(fill="navyblue")+
  labs(x="Number of registered death per capita", y="Number of countries")

# Summary statistics of x variable
library(moments)

sum_death <- df %>% summarise(
  variable='death_per_capita',
  n   = n(),
  Mean     = mean(death_per_capita),
  Median   = median(death_per_capita),
  Std      = sd(death_per_capita),
  IQ_range = IQR(death_per_capita), 
  Min      = min(death_per_capita),
  Max      = max(death_per_capita),
  Skew     = skewness(death_per_capita))

# Checking the distribution of y variable
ggplot(data = df , aes(x = case_per_capita)) +
  geom_histogram(fill="navyblue")+
  labs(x="Number of registered case per capita", y="Number of countries")

# Summary statistics of x variable
sum_case <- df %>% summarise(
  variable='case_per_capita',
  n   = n(),
  Mean     = mean(case_per_capita),
  Median   = median(case_per_capita),
  Std      = sd(case_per_capita),
  IQ_range = IQR(case_per_capita), 
  Min      = min(case_per_capita),
  Max      = max(case_per_capita),
  Skew     = skewness(case_per_capita))

# Both of the variables are skewed with a long right tale. 
```
## Choosing a model

As the dependent variable and the explanatory variable both skewed with long right tale and does not contain negative numbers, the log transformation could be applied on the values. We compared four different models: 
* level-level model
* log-level model
* level-log model
* log-log model

On the plots, we can observe that the log-log model makes the association close to linear between the both variable. This model shows us, that how many percent higher number of death per capita associated with 1% higher number of cases per capita by country, which is a meaningful interpretation. Furhtermore, both of the variables are skewed with a long right tale, so the log transformation balance the distribution for the further analysis. So, the log-log regression is going to be analysed in the further analysis. 

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', fig.show='hide'}
##
# Model: death_per_capita = alpha + beta*case_per_capita 
# Checking the possible different ln transformations 

# death_per_capita - case_per_capita: level-level model without scaling
ggplot( df , aes(x = case_per_capita, y = death_per_capita)) +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of registered case per capita", y = "Number of registered death per capita") 

# death_per_capita - case_per_capita: log-level model 
require(scales)

ggplot( df , aes(x = case_per_capita, y = death_per_capita ))  +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of registered case per capita, ln scale", y = "Number of registered death per capita") +
  scale_y_continuous( trans = log_trans() )

# death_per_capita - case_per_capita: level-log model 
ggplot( df , aes(x = case_per_capita, y = death_per_capita ))  +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of registered case per capita", y = "Number of registered death per capita, ln scale") +
  scale_x_continuous( trans = log_trans() )


```
```{r log-log, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
require(scales)

ggplot( df , aes(x = case_per_capita, y = death_per_capita ))  +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of registered case per capita, ln scale", y = "Number of registered death per capita, ln scale") +
  scale_x_continuous( trans = log_trans() )+
  scale_y_continuous( trans = log_trans() )


# Conclusion: taking log-log makes the association close to linear
# Substantive: van-e értelme a kérdésnek?
#         - variables are affected in multiplicative ways
#         - variables are ratios
# Statistical: 
#         - both of the variables are skewed with long right tale --> take log
```


## Regression

In the analysis, we make five regression:

* Simple linear regression (log-log):
  + reg1: ln_death_per_capita = alpha + beta * ln_case_per_capita

* Quadratic linear regression:
  + reg2: ln_death_per_capita  = alpha + beta_1 * ln_case_per_capita + beta_2 * ln_case_per_capita^2

* Cubic liear regression:
  + reg3: ln_death_per_capita  = alpha + beta_1 * ln_case_per_capita + beta_2 * ln_case_per_capita^2 + beta_3 * ln_case_per_capita^3

* Piecewise linear spline regression
  + reg4: ln_death_per_capita = alpha + beta_1 * ln_case_per_capita * 1(case_per_capita < 0.00091) + beta_2 * ln_case_per_capita * 1(case_per_capita >= 0.00091)

* Weighted linear regression, using population as weights
  + reg5: ln_death_per_capita = alpha + beta * ln_case_per_capita, weights: population
  
For the further analysis, we pick the most reliable model. Based on the adjusted R-squared, it is the weighted linear regression model, which uses population as weights. In this case, the R-squared is 0.89, which could be accepted as a relatively high R-squared value. 

The result of the weighted linear regression model is ln_death_per_capita = -4.25 + 0.91 * ln_case_per_capita, which means that the number of death is 0.91% higher on average for observations with one percent higher number of cases due to COVID.

```{r, message=FALSE, warning=FALSE, echo=FALSE, results='hide', fig.show='hide'}
# Take log of death_per_capita and log of case_per_capita
df <- df %>% mutate( ln_death_per_capita = log( death_per_capita ),
                     ln_case_per_capita = log( case_per_capita ) )

## 
# Making models

# Simple linear regression (log-log):
# reg1: ln_death_per_capita = alpha + beta * ln_case_per_capita

# Quadratic linear regression:
# reg2: ln_death_per_capita  = alpha + beta_1 * ln_case_per_capita + beta_2 * ln_case_per_capita^2

# Cubic liear regression:
# reg3: ln_death_per_capita  = alpha + beta_1 * ln_case_per_capita + beta_2 * ln_case_per_capita^2 + beta_3 * ln_case_per_capita^3

# Piecewise linear spline regression
# reg4: ln_death_per_capita = alpha + beta_1 * ln_case_per_capita * 1(case_per_capita < 0.00091) + beta_2 * ln_case_per_capita * 1(case_per_capita >= 0.00091)

# Weighted linear regression, using population as weights
# reg5: ln_death_per_capita = alpha + beta * ln_case_per_capita, weights: population

# Add powers of the variable(s) to the dataframe
df <- df %>% mutate( ln_case_per_capita_sq = ln_case_per_capita^2 ,
                     ln_case_per_capita_cb = ln_case_per_capita^3)

# REGRESSION

# Removing infinite values
df <- df %>% 
  filter_all(all_vars(!is.infinite(.)))

# Reg0 - built in regression in R
reg_b <- lm( ln_death_per_capita ~ ln_case_per_capita , data = df )
reg_b
summary( reg_b )
## In this case, only homoskedastic SEs, which is not robust

# Reg1 - log-log regression with lm_robust
library(estimatr)

reg1 <- lm_robust( ln_death_per_capita ~ ln_case_per_capita , data = df , se_type = "HC2" )

summary( reg1 )

reg1_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg2 - quadratic linear regression
reg2 <- lm_robust( ln_death_per_capita ~ ln_case_per_capita + ln_case_per_capita_sq , data = df )

summary( reg2 )

reg2_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( formula = y ~ poly(x,2) , method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg3 - cubic linear regression
reg3 <- lm_robust( ln_death_per_capita ~ ln_case_per_capita + ln_case_per_capita_sq + ln_case_per_capita_cb , data = df )

summary (reg3)

reg3_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( formula = y ~ poly(x,3) , method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg4 - piecewiese linear regression
library(lspline)

cutoff_1 <- 0.0005
cutoff_2 <- 0.0096
cutoff_ln_1 <- log(cutoff_1)
cutoff_ln_2 <- log(cutoff_2)
reg4 <- lm_robust(ln_death_per_capita ~ lspline( ln_case_per_capita , cutoff_ln_1 , cutoff_ln_2 ), data = df )
summary( reg4 )
reg4_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( formula = y ~ lspline(x,cutoff_ln_1 , cutoff_ln_2) , method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg5 - weighted linear regression
reg5 <- lm_robust(ln_death_per_capita ~ ln_case_per_capita, data = df , weights = population)
summary( reg5 )
reg5_plot <- ggplot(data = df, aes(x = ln_case_per_capita, y = ln_death_per_capita)) +
  geom_point(data = df, aes(size=population),  color = 'blue', shape = 16, alpha = 0.6,  show.legend=F) +
  theme_classic()+
  geom_smooth(aes(weight = population), method = "lm", color='red')+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Comparing the plots


```
```{r model comparison, message=FALSE, warning=FALSE, echo=FALSE}
library(patchwork)
reg1_plot+reg2_plot+reg3_plot+reg4_plot+reg5_plot
```



## Hypothesis testing

In the hypothesis testing, we test if there is a significant linear relationship between the number of death and number of cases on a 0.05 level of significance. 

* H0: beta = 0, or the slope of the regression line is zero
* H1: beat != 0, or the slope of the regression line is not equal to zero. If the relationship between the death and cases is significant, the slope will not equal to zero. 

The p-value of the model is 2.2e-16, which is less than the significance level (0.05), we cannot accept the null hypothesis.
```{r, message=FALSE, warning=FALSE, echo=FALSE}
######
## Hypothesis testing
#####
#       H0: beta=0, 
#       HA: beta != 0
library(car)
linearHypothesis( reg5 , "ln_case_per_capita=0")
```

## Residual analysis

Based on the residual analysis, we can observe, which countries lost and save relatively the most people. The residual analysis shows, which observations are the farest from the predicted value. Countries, where there are less death due to COVID-19 compared to the number of cases, are shown Table 1. While those countries, where there are more death due to COVID-19 compared to the number of cases, are shown Table 2.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
######
# Residual analysis.
library(magrittr)
library(dplyr)

# Getting the predicted y values from the model
df$reg5_y_pred <- reg5$fitted.values

# Calculating the errors of the model
df$reg5_res <- df$ln_death_per_capita - df$reg5_y_pred 

# Finding countries with largest negative errors (who saved relatively the most people)
library(knitr)
kable(df %>% top_n( -5 , reg5_res ) %>% 
  select( country , ln_death_per_capita , reg5_y_pred , reg5_res ), caption="The TOP5 countries, who saved relatively the most people")

# Finding countries with largest positive errors (who lost relatively the most people)
kable(df %>% top_n( 5 , reg5_res ) %>% 
  select( country , ln_death_per_capita , reg5_y_pred , reg5_res ), caption = "The TOP5 countries, who saved relatively the most people")
```


# Appendix

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(knitr)
kable(summary( df ) , caption = "Descriptive statistics of the variables")

# Checking the distribution of x variable
ggplot(data = df , aes(x = death_per_capita)) +
  geom_histogram(fill="navyblue")+
  labs(x="Number of registered death per capita", y="Number of countries")

# Summary statistics of x variable
library(moments)

sum_death <- df %>% summarise(
  variable='death_per_capita',
  n   = n(),
  Mean     = mean(death_per_capita),
  Median   = median(death_per_capita),
  Std      = sd(death_per_capita),
  IQ_range = IQR(death_per_capita), 
  Min      = min(death_per_capita),
  Max      = max(death_per_capita),
  Skew     = skewness(death_per_capita))

# Checking the distribution of y variable
ggplot(data = df , aes(x = case_per_capita)) +
  geom_histogram(fill="navyblue")+
  labs(x="Number of registered case per capita", y="Number of countries")

# Summary statistics of x variable
sum_case <- df %>% summarise(
  variable='case_per_capita',
  n   = n(),
  Mean     = mean(case_per_capita),
  Median   = median(case_per_capita),
  Std      = sd(case_per_capita),
  IQ_range = IQR(case_per_capita), 
  Min      = min(case_per_capita),
  Max      = max(case_per_capita),
  Skew     = skewness(case_per_capita))

##
# Model: death_per_capita = alpha + beta*case_per_capita 
# Checking the possible different ln transformations 

# death_per_capita - case_per_capita: level-level model without scaling
ggplot( df , aes(x = case_per_capita, y = death_per_capita)) +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of registered case per capita", y = "Number of registered death per capita") 

# death_per_capita - case_per_capita: log-level model 
require(scales)

ggplot( df , aes(x = case_per_capita, y = death_per_capita ))  +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of registered case per capita, ln scale", y = "Number of registered death per capita") +
  scale_y_continuous( trans = log_trans() )

# death_per_capita - case_per_capita: level-log model 
ggplot( df , aes(x = case_per_capita, y = death_per_capita ))  +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of registered case per capita", y = "Number of registered death per capita, ln scale") +
  scale_x_continuous( trans = log_trans() )

# REGRESSION

# Removing infinite values
df <- df %>% 
  filter_all(all_vars(!is.infinite(.)))

# Reg0 - built in regression in R
reg_b <- lm( ln_death_per_capita ~ ln_case_per_capita , data = df )
reg_b
summary( reg_b )
## In this case, only homoskedastic SEs, which is not robust

# Reg1 - log-log regression with lm_robust
library(estimatr)

reg1 <- lm_robust( ln_death_per_capita ~ ln_case_per_capita , data = df , se_type = "HC2" )

summary( reg1 )

reg1_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg2 - quadratic linear regression
reg2 <- lm_robust( ln_death_per_capita ~ ln_case_per_capita + ln_case_per_capita_sq , data = df )

summary( reg2 )

reg2_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( formula = y ~ poly(x,2) , method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg3 - cubic linear regression
reg3 <- lm_robust( ln_death_per_capita ~ ln_case_per_capita + ln_case_per_capita_sq + ln_case_per_capita_cb , data = df )

summary (reg3)

reg3_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( formula = y ~ poly(x,3) , method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg4 - piecewiese linear regression
library(lspline)

cutoff_1 <- 0.0005
cutoff_2 <- 0.0096
cutoff_ln_1 <- log(cutoff_1)
cutoff_ln_2 <- log(cutoff_2)
reg4 <- lm_robust(ln_death_per_capita ~ lspline( ln_case_per_capita , cutoff_ln_1 , cutoff_ln_2 ), data = df )
summary( reg4 )
reg4_plot <- ggplot( data = df, aes( x = ln_case_per_capita, y = ln_death_per_capita ) ) + 
  geom_point( color='blue') +
  geom_smooth( formula = y ~ lspline(x,cutoff_ln_1 , cutoff_ln_2) , method = lm , color = 'red' )+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

# Reg5 - weighted linear regression
reg5 <- lm_robust(ln_death_per_capita ~ ln_case_per_capita, data = df , weights = population)
summary( reg5 )
reg5_plot <- ggplot(data = df, aes(x = ln_case_per_capita, y = ln_death_per_capita)) +
  geom_point(data = df, aes(size=population),  color = 'blue', shape = 16, alpha = 0.6,  show.legend=F) +
  theme_classic()+
  geom_smooth(aes(weight = population), method = "lm", color='red')+
  labs(x = "ln(Number of registered case per capita)", y = "ln(Number of registered death per capita)")

```

